The Decreasing Orbital Period of WASP-4\lowercase{b}

### 2018-12-06 paper/ms.tex
minor updates

%-------------------------------------
% TESS Mission Architects:
% These authors should be listed in this order
% see https://spacebook.mit.edu/pages/viewpage.action?pageId=24543276
%-------------------------------------
%\\ \textcolor{red}{6 TESS Mission Architects:}
\author{G. R. Ricker} % grr@space.mit.edu WAITING
\affiliation{ Department of Physics and Kavli Institute for Astrophysics
and Space Research, Massachusetts Institute of Technology, Cambridge, MA
02139, USA }
\author[0000-0001-6763-6562]{R. Vanderspek} % roland@space.mit.edu WAITING
\affiliation{ Department of Physics and Kavli Institute for Astrophysics
and Space Research, Massachusetts Institute of Technology, Cambridge, MA
02139, USA }
\author[0000-0001-9911-7388]{D. W.~Latham} % dlatham@cfa.harvard.edu WAITING
\affiliation{ Harvard-Smithsonian Center for Astrophysics, 60 Garden
Street, Cambridge, MA 02138, USA }
\author{S. Seager} % seager@mit.edu WAITING
\affiliation{ Department of Earth, Atmospheric, and Planetary Sciences,
Massachusetts Institute of Technology, Cambridge, MA 02139, USA }
\author[0000-0002-4715-9460]{J. M.~Jenkins} % jon.jenkins@nasa.gov WAITING
\affiliation{ NASA Ames Research Center, Moffett Field, CA 94035, USA }
%-------------------------------------
% 3 representatives of each of SPOC, POC, TSO, for a total of 9. 
%These 9 authors should be listed in alphabetical order
%-------------------------------------
%\\ \textcolor{red}{3 TSO coauthors:}
\author{TSO1}
\author{TSO2}
\author{TSO3}
%\\ \textcolor{red}{3 SPOC coauthors:}
\author{J. D. Twicken} % Joseph.Twicken@nasa.gov WAITING
\affiliation{ NASA Ames Research Center, Moffett Field, CA 94035, USA }
\affiliation{ SETI Institute, Mountain View, CA, 94043}
\author{B. Wohler} % bill.wohler@nasa.gov WAITING
\affiliation{ NASA Ames Research Center, Moffett Field, CA 94035, USA }
\affiliation{ SETI Institute, Mountain View, CA, 94043}
\author{D. A. Caldwell} % douglas.Caldwell@nasa.gov % WAITING
\affiliation{ NASA Ames Research Center, Moffett Field, CA 94035, USA }
\affiliation{ SETI Institute, Mountain View, CA, 94043}
%\\ \textcolor{red}{3 POC coauthors:}
\author{POC1}
\author{POC2}
\author{POC3}

@@

Figure~\ref{fig:lightcurves} shows the light curves, best-fit transit
models, and residuals.  Table~1 reports the mid-transit times and
their uncertainties.

Taking the mean and standard deviation of the measured planet to star
radius ratio, we find $R_{\rm p}/R_\star=0.1538\pm0.0013$, in
agreement with previous determinations
\citep{wilson_wasp-4b_2008,gillon_improved_2009,winn_transit_2009,southworth_high-precision_2009}.

Binning the residuals to 1-hour windows and taking the median absolute
deviation, the residual lightcurves have a MAD of $594\,{\rm
ppm}\,{\rm hr}^{1/2}$.  WASP-4 has a TESS-band magnitude of $T=11.78$
\citep{stassun_TIC_2018}, corresponding to a predicted photon-counting
noise\footnote{\url{github.com/lgbouma/tnm}} of $410\,{\rm ppm}\,{\rm
hr}^{1/2}$.  At $T=11.78$, with a 9-pixel aperture read noise is
expected to contribute an additional $202\,{\rm ppm}\,{\rm hr}^{1/2}$ in
quadrature.  The predicted zodiacal background contribution is
$673\,{\rm ppm}\,{\rm hr}^{1/2}$, based on the
\citet{winn_photonflux_2013} model, which \citet{Sullivan_2015}
used in planet detection simulations.  The predicted level
of zodiacal noise seems to have been an overestimate, given the observed
RMS of WASP-4.  A more detailed assessment of TESS's photometric
performance is beyond the scope of this work.

@@

%TODO: final numbers
The difference in $\chi^2$ between the linear and quadratic fit
corresponds to $p \approx 10^{-23}$. %In other words, assuming the linear
%model is true, there is a $10^{-23}$ probability that simply by chance
%we would see data that favors the quadratic model more strongly than the
%observations.
Clearly, it is more likely that some systematic error could be
affecting one of the most important sets of times: either the earliest
times, the Huitson et al.\ Gemini times (epoch numbers 44, 322, 591,
and 851), or the TESS times.  We have taken every precaution against
systematic offsets in the TESS dataset
(Appendix~\ref{sec:verify_tess}), and Huitson et al.\ stand by their
reported times (Huitson, priv.  comm.).  Since the linear model
provides a poor fit to the data, we discard it from further
consideration.

@@

Let us assume further that the ``constant phase lag'' model for tidal
interaction in a binary applies \citep{zahn_tidal_1977}.  The fluid
response is parametrized by a constant modified\footnote{For stars,
$k_\star \sim \mathcal{O}(10^{-2})$, so it is important to explicitly
distinguish $Q_\star'$ from $Q_\star$ \citep[{\it
e.g.},][]{schwarzschild_structure_1958}.} quality factor, $Q_\star' =
3 Q_\star / (2k_\star)$.  Here $k_\star$ is the stellar Love number,
which is smaller when the star's density distribution is more
centrally concentrated. $Q_\star$ is the ratio of the energy stored in
the equilibrium deformation of the star, divided by the energy
lost to heat per tidal period \citep[{\it e.g.},][]{goldreich_q_1966}.
A larger $Q_\star$ implies less efficient dissipiation.  Though the
fluid mechanics responsible for the dissipation should depend on the
amplitude and frequency of the tidal perturbation
\citep[][Section~3.3]{ogilvie_tidal_2014}, we will treat $Q_\star'$ as
a constant.  This should not be taken literally; $Q_\star'$ may in
fact change sharply with varying tidal frequency
\citep{penev_empirical_2018}.

If the planet's spin and orbit are synchronized, the star is slowly
rotating, and the eccentricity is small, then the semi-major axis and
eccentricity evolve as
\citep[Appendix B of][]{metzger_optical_2012}
\begin{align}
  \frac{1}{\tau_{\rm e}} &=
  \frac{|\dot{e}|}{e} =
    \frac{63 \pi } {2 Q_{\rm p}' }
    \left( \frac{R_{\rm p}}{a} \right)^5
    \left( \frac{M_\star}{M_{\rm p}} \right)
    \left( \frac{1}{P} \right)
  \label{eq:de_dt}
  \\
  \frac{1}{\tau_{\rm a}} &=
  \frac{|\dot{a}|}{a} =
    \frac{9 \pi } {Q_\star' }
    \left( \frac{R_\star}{a} \right)^5
    \left( \frac{M_{\rm p}}{M_\star} \right)
    \left( \frac{1}{P} \right).
  \label{eq:da_dt}
\end{align}
The orbital period evolves as
\begin{equation}
\label{eq:dP_dt}
  \dot{P} = -\frac{27\pi}{2 Q_\star'}
            \left(\frac{M_{\rm p}}{M_\star}\right)
            \left(\frac{R_\star}{a}\right)^5.
\end{equation}
Inserting our observed value of $\dot{P}$, the current dissipation
rate corresponds to a modified quality factor for WASP-4 of
\begin{equation}
	Q_\star' \approx 3.3\times10^4. 
\end{equation}
This is rather small.  Jupiter has $Q_{\rm Jup}' \approx 1.4 \times
10^5$, based on the motion of the Galilean moons
\citep{lainey_strong_2009}.  For stars, different $Q_\star'$ values have
been estimated in different contexts.  Studies of the observed
eccentricity distribution in stellar binaries have yielded $10^5 \gtrsim
Q_\star' \gtrsim 10^{7}$ \citep[{\it
e.g.},][]{meibom_robust_2005,belczynski_compact_2008,
geller_direct_2013,milliman_wiyn_2014}.  However the dissipation rates
may be different in hot Jupiter systems, since the tidal forcing
frequencies and amplitudes are different.  In fact, for the observed
population of hot Jupiters, \citet{penev_constraining_2012} argued that
$Q_\star' > 10^7$ would be required at some phase of hot Jupiter
evolution at the 99\% level, or else the observed population would not
exist (though \citealt{birkby_wts-2_2014} point out more efficient
dissipation would be allowed if the hot Jupiters arrived later).
Developing a generative model for the orbital separation distribution of
the hot Jupiter population, \citet{cameron_hierarchical_2018} similarly
found $10^7 \gtrsim Q_\star' \gtrsim 10^8$.

>>>>

%-------------------------------------
% TESS Mission Architects:
% These authors should be listed in this order
% see https://spacebook.mit.edu/pages/viewpage.action?pageId=24543276
%-------------------------------------
%\\ \textcolor{red}{6 TESS Mission Architects:}
\author{G. R. Ricker} % grr@space.mit.edu WAITING
\affiliation{ Department of Physics and Kavli Institute for Astrophysics
and Space Research, Massachusetts Institute of Technology, Cambridge, MA
02139, USA }
\author[0000-0001-6763-6562]{R. Vanderspek} % roland@space.mit.edu WAITING
\affiliation{ Department of Physics and Kavli Institute for Astrophysics
and Space Research, Massachusetts Institute of Technology, Cambridge, MA
02139, USA }
\author[0000-0001-9911-7388]{D. W.~Latham} % dlatham@cfa.harvard.edu WAITING
\affiliation{ Harvard-Smithsonian Center for Astrophysics, 60 Garden
Street, Cambridge, MA 02138, USA }
\author{S. Seager} % seager@mit.edu WAITING
\affiliation{ Department of Earth, Atmospheric, and Planetary Sciences,
Massachusetts Institute of Technology, Cambridge, MA 02139, USA }
\author[0000-0002-4715-9460]{J. M.~Jenkins} % jon.jenkins@nasa.gov WAITING
\affiliation{ NASA Ames Research Center, Moffett Field, CA 94035, USA }
%-------------------------------------
% 3 representatives of each of SPOC, POC, TSO, for a total of 9. 
%These 9 authors should be listed in alphabetical order
%-------------------------------------
% 3 TSO COAUTHORS
%
\author[0000-0002-3321-4924]{Z. Berta-Thompson} % Zachory.BertaThompson@Colorado.EDU WAITING
\affiliation{Department of Astrophysical and Planetary Sciences,
University of Colorado, Boulder, CO 80309, USA}
%
\author[0000-0001-8812-0565]{J. Rodriguez} % rodriguez.jr.joey@gmail.com WAITING
\affiliation{Harvard-Smithsonian Center for Astrophysics, 60 Garden
Street, Cambridge, MA 02138, USA }
%
\author[0000-0001-8020-7121]{K. Col\'on} % knicole.colon@nasa.gov WAITING
\affiliation{NASA Goddard Space Flight Center, Exoplanets and Stellar
Astrophysics Laboratory (Code 667), Greenbelt, MD 20771, USA}
% 3 SPOC COAUTHORS
%
\author[0000-0002-6778-7552]{J. D. Twicken} % Joseph.Twicken@nasa.gov WAITING
\affiliation{ NASA Ames Research Center, Moffett Field, CA 94035, USA }
\affiliation{ SETI Institute, Mountain View, CA, 94043}
%
\author{B. Wohler} % bill.wohler@nasa.gov WAITING
\affiliation{ NASA Ames Research Center, Moffett Field, CA 94035, USA }
\affiliation{ SETI Institute, Mountain View, CA, 94043}
%
\author[0000-0003-1963-9616]{D. A. Caldwell} % douglas.Caldwell@nasa.gov % WAITING
\affiliation{ NASA Ames Research Center, Moffett Field, CA 94035, USA }
\affiliation{ SETI Institute, Mountain View, CA, 94043}
%
%\\ \textcolor{red}{3 POC coauthors:}
\author{POC1}
\author{POC2}
\author{POC3}

@@

Figure~\ref{fig:lightcurves} shows the light curves, best-fit transit
models, and residuals.  Table~1 reports the mid-transit times and
their uncertainties.
Taking the mean and standard deviation of the measured planet to star
radius ratio, we find $R_{\rm p}/R_\star=0.1538\pm0.0013$, in
agreement with previous determinations
\citep{wilson_wasp-4b_2008,gillon_improved_2009,winn_transit_2009,southworth_high-precision_2009}.
Binning the residuals to 1-hour windows and taking the median absolute
deviation, the residual lightcurves have a MAD of $594\,{\rm
ppm}\,{\rm hr}^{1/2}$.  WASP-4 has a TESS-band magnitude of $T=11.78$
\citep{stassun_TIC_2018}, corresponding to a predicted photon-counting
noise\footnote{\url{github.com/lgbouma/tnm}} of $410\,{\rm ppm}\,{\rm
hr}^{1/2}$.  At $T=11.78$, with a 9-pixel aperture read noise is
expected to contribute an additional $202\,{\rm ppm}\,{\rm hr}^{1/2}$ in
quadrature.  The predicted zodiacal background contribution is
$673\,{\rm ppm}\,{\rm hr}^{1/2}$, based on the
\citet{winn_photonflux_2013} model, which \citet{Sullivan_2015}
used in planet detection simulations.  The predicted level
of zodiacal noise seems to have been an overestimate, given the observed
RMS of WASP-4.  A more detailed assessment of TESS's photometric
performance is beyond the scope of this work.

@@

%TODO: final numbers
The difference in $\chi^2$ between the linear and quadratic fit
corresponds to $p \approx 10^{-23}$.
%In other words, assuming the linear model is true, there is a
%$10^{-23}$ probability that simply by chance we would see data that
%favors the quadratic model more strongly than the observations.
Clearly, it is more likely that some systematic error could be
affecting one of the most important sets of times: either the earliest
times, the Huitson et al.\ Gemini times (epoch numbers 44, 322, 591,
and 851), or the TESS times.  We have taken every precaution against
systematic offsets in the TESS dataset
(Appendix~\ref{sec:verify_tess}). Huitson et al.\ stand by their
reported times (J.-M. D\'esert, priv.\
comm.)\footnote{\citealt{huitson_gemini_2017} corrected the times to
be at mid-exposure. They also analyzed other hot Jupiters, which all
showed mid-transit times as expected from a constant ephemeris.}.
Since the linear model provides a poor fit to the data, we discard it
from further consideration.

@@

Let us assume further that the ``constant phase lag'' model for tidal
interaction in a binary applies \citep{zahn_tidal_1977}.  The fluid
response is parametrized by a modified\footnote{For stars,
$k_\star \sim \mathcal{O}(10^{-2})$, so it is important to explicitly
distinguish $Q_\star'$ from $Q_\star$ \citep[{\it
e.g.},][]{schwarzschild_structure_1958}.} quality factor, $Q_\star' =
3 Q_\star / (2k_\star)$.  Here $k_\star$ is the stellar Love number,
which is smaller when the star's density distribution is more
centrally concentrated. $Q_\star$ is the ratio of the energy stored in
the equilibrium deformation of the star, divided by the energy
lost to heat per tidal period \citep[{\it e.g.},][]{goldreich_q_1966}.
A larger $Q_\star$ implies less efficient dissipiation.  Though the
fluid mechanics responsible for the dissipation should depend on the
amplitude and frequency of the tidal perturbation
\citep[][Section~3.3]{ogilvie_tidal_2014}, we will treat $Q_\star'$ as
a constant.  This should not be taken literally; $Q_\star'$ may in
fact change sharply with varying tidal frequency
\citep{penev_empirical_2018}.

If the planet's spin and orbit are synchronized, the star is slowly
rotating, and the eccentricity is small, then the semi-major axis and
eccentricity evolve as
\citep[Appendix B of][]{metzger_optical_2012}
\begin{align}
  \frac{1}{\tau_{\rm e}} &=
  \frac{|\dot{e}|}{e} =
    \frac{63 \pi } {2 Q_{\rm p}' }
    \left( \frac{R_{\rm p}}{a} \right)^5
    \left( \frac{M_\star}{M_{\rm p}} \right)
    \left( \frac{1}{P} \right)
  \label{eq:de_dt}
  \\
  \frac{1}{\tau_{\rm a}} &=
  \frac{|\dot{a}|}{a} =
    \frac{9 \pi } {Q_\star' }
    \left( \frac{R_\star}{a} \right)^5
    \left( \frac{M_{\rm p}}{M_\star} \right)
    \left( \frac{1}{P} \right).
  \label{eq:da_dt}
\end{align}
The orbital period evolves as
\begin{equation}
\label{eq:dP_dt}
  \dot{P} = -\frac{27\pi}{2 Q_\star'}
            \left(\frac{M_{\rm p}}{M_\star}\right)
            \left(\frac{R_\star}{a}\right)^5.
\end{equation}
Using our observed value of $\dot{P}$, the current modified quality
factor of WASP-4 would be
\begin{equation}
	Q_\star' \approx 3.3\times10^4. 
\end{equation}
This is rather small.  Jupiter has $Q_{\rm Jup}' \approx 1.4 \times
10^5$, based on the motion of the Galilean moons
\citep{lainey_strong_2009}.  For stars, different $Q_\star'$ values have
been estimated in different contexts.  Studies of the 
eccentricity distribution of stellar binaries have yielded $10^5 \gtrsim
Q_\star' \gtrsim 10^{7}$ \citep[{\it
e.g.},][]{meibom_robust_2005,belczynski_compact_2008,
geller_direct_2013,milliman_wiyn_2014}.  However the dissipation rates
may be different in hot Jupiter systems, since the tidal forcing
frequencies and amplitudes are different.  In fact, for the observed
population of hot Jupiters, \citet{penev_constraining_2012} argued that
$Q_\star' > 10^7$ would be required at some phase of hot Jupiter
evolution at the 99\% level, or else the observed population would not
exist (though \citealt{birkby_wts-2_2014} point out more efficient
dissipation would be allowed if the hot Jupiters arrived later).
Developing a generative model for the orbital separation distribution of
the hot Jupiter population, \citet{cameron_hierarchical_2018} similarly
found $10^7 \gtrsim Q_\star' \gtrsim 10^8$.
